A continental-scale analysis finds generalists, with large clutch size and residual brain mass, are the most urban-tolerant of birds

### 2018-05-21 ms.Rmd
more minor tweaks to manuscript

We filtered the bird observations [@callaghan2017assessing; @la2014role] by only including observations which were on complete checklists and which followed these protocols: travelling, random, stationary, area, or BirdLife Australia protocols [@barrett2003new]. For further information on eBird protocols see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types. Further, we filtered the checklists to those with a distance travelled of < 5 km or an area covered which was < 500 Ha. Any checklists which were shared among multiple individuals were subsampled to include only one checklist in order to avoid duplicate results. After this sampling, we were left with 637,482 eligible checklists throughout Australia.

Seabirds (gannets, petrels, shearwaters, etc.) were omitted from the analyses as we did not expect any correlation with seabirds using urban areas under normal circumstances. Only species which included > 100 observations in the eBird database over the specified time-frame, and that met our criteria were considered for further analyses. This left us with 580 potential species from the eBird dataset which had > 100 observations on eligible checklists in the specified time frame, which was a substantial proportion of the ~730 bird species which are resident in Australia.

Given the diverse datasets (see below) used in this analysis, there were differing taxonomic authorities followed (see Appendix 2 for elaboration).

## Urbanization index
We used VIIRS night-time lights [@elvidge2017viirs] as a proxy of a bird's association with urban habitat. The VIIRS night-time lights measures the electric lighting present on Earth's surface which is generally associated with human settlement. There are also a significant number of steps (e.g., filtering out background noise, degraded data, and other light source contamination) which are used to ensure the validity of the data and its association with human settlement [@elvidge2017viirs]. For the 637,482 checklists which met our criteria above, we used the checklists' associated spatial coordinates and Google Earth Engine [@gorelick2017google] to assign the average radiance value calculated from the VIIRS layer. For each checklist, the average radiance was calculated within a 5 km buffer, in order to match the spatial scale of the checklists included and to minimize any potential bias in eBird sampling protocols. Any radiance values which were negative were forced to 0.00001 [*sensu* @ou2015evaluation].

Each checklist's average radiance value was then passed to every bird on that checklist, which provided every species with a distribution of average radiance values (Fig. 1). The median for each species' given distribution of average radiance values served as our response variable in our analyses.

@@

Before modelling, the predictor variables were investigated for multi-collinearity. We used the corrplot package [@corrplot2017] to show correlation among continuous variables (Supplementary Figure 1), and used generalized variance inflation factors [@fox1992generalized], from the car package [@fox2011companion], which accounts for the degrees of freedom for a particular term to show minimal correlation amongst each other (Supplementary Table 1).

@@

### Phylogenetic analysis

For the phylogenetic portion of our analyses, we used the phylogenetic trees provided by Jetz et al. [@jetz2012global]. This phylogeny represents the first completely mapped phylogeny of 9,993 birds [@jetz2012global]. In order to assess whether there was a phylogenetic signal [@revell2008phylogenetic; @munkemuller2012measure] of the urbanization index and the continuous predictor variables used in our study, we used the phylosignal package [@keck2016phylosignal] in the R computing environment [@r2017coreteam]. This was also conducted for each of the continuous response variables in order to assess the relative contribution of phylogeny on the predictor variables.

@@

# RESULTS
A total of 316,306 checklists resulting in 5,944,819 observations of 477 different species were included in analyses. The least often recorded species was Black-breasted Buttonquail (N=104) while the most recorded species was Australian Magpie (N=158,615). The mean number of observations for a species was 10,250. Plots of all 477 species' distributions of average radiance values are presented in Appendix 3. A complete list of the 477 species, in addition to their traits, number of observations, and their urbanization indices can be found in Appendix 4.

## Urbanization index
For the 477 species included in the models the mean $\pm$ SD urbanization index was 0.91 $\pm$ 1.76, with a range from 0.008 (Black-backed Butcherbird) to 11.85 (Powerful Owl; Fig. 2).

### Phylogenetic signal
There was a distinct phylogenetic signal in our urbanization index measure (Fig. 3; Fig. S2; Table 1), as it was significantly different than random for 4 out of the 5 indices considered. Additionally, there was strong phylogenetic signal for potential predictor traits, especially body size and brain residual (Table 1). The result for body size ($k=3.3$) is among the strongest ever recorded measures of phylogenetic signal in the literature, and for $k$ the statistic suggests much stronger signal than expected from Brownian motion. We generally found weaker signal for behavioral traits compared with morphological traits. For these traits, the estimates of phylogenetic signal were weaker than Brownian motion but significant in most cases (Table 1).

@@

# DISCUSSION
Niche breadth, as measured by different types of generalism, was a significantly important life-history characteristic associated with the most urban-tolerant species in our analyses. This result was robust across non-phylogenetic and phylogenetic models, model-averaging approaches, and across phylogenetic uncertainty (Fig. S3). Conversely, specialized feeding strategies were negatively associated with urbanization: granivorous and insectivorous specialists were much less likely to be found in urban environments. These results suggest that the process of urbanization is, in general, operating to exclude specialist species [@sorace2009distribution]. Given specialist species are known to have strong negative responses to landscape fragmentation and disturbance [@devictor2008distribution], sustainable urban development with a goal of maintaining biological diversity in urban areas should focus on providing intact vegetation [@sushinsky2013should] connected by corridors [@litteral2017role]. A species-specific approach, which assesses particular resources necessary for specialist species, for example focusing on foraging [@shochat2004linking] and nesting [@fernandez2001habitat] requirements, is likely necessary. We recommend that a focal point of avian biodiversity conservation should be on increasing insect abundance and diversity in urban environments [@baldock2015uk], especially important given the recent noted collapse of insects in other parts of the world [@hallmann2017more].

The non-random distribution of the urbanization index across the phylogeny (Table 1) is consistent with the observation of reduced phylogenetic diversity in urban areas [@la2018phylogenetic], suggesting that particular phylogenetic constraints allow certain species to remain tolerant of urban environments [@morelli2016evidence; @ibanez2017global; @sol2017urbanisation]. This finding of a strong phylogenetic signal contrasts with previous studies [@evans2011makes], but this is likely a result of increased statistical power from our larger sample size: 477 species in this study compared with 55 species in Evans *et al.* [-@evans2011makes]. Unsurprisingly, our larger sample size found a non-random pattern in urbanization across the bird phylogeny and we further demonstrated the importance of treating species as non-independent by considering phylogeny in the context of the linear models. For example, some predictor variables became less important in the phylogenetic models compared with the non-phylogenetic models, including some feeding types (i.e., insect specialists) and habitat preferences (i.e., agricultural). For these variables, it is likely that there may be particular clades which include a number of species that are associated with both that particular trait and degree of urbanization [see @felsenstein1985confidence]. Conversely, some of the most important predictor variables (e.g., clutch size and feeding habitat generalism; Figures 4 and 5) showed a stronger relationship (i.e., increased parameter estimates) with urbanization when phylogeny was accounted for, and this affect was robust relative to a consideration of phylogenetic uncertainty (Fig. S3). Furthermore, residual brain mass (and body size to a lesser extent) was significantly important within the phylogenetic models, but not in the non-phylogenetic models, suggesting that within a clade the species with the largest residual brain mass [@maklakov2011brains] and largest body size [@major2010museum] are most suited to surviving in urban areas. While some studies have found a lack of evidence that residual brain size is an important predictor in urban-tolerant species [@kark2007living; @evans2011makes], our results suggest that residual brain size is indeed heavily correlated with urban-tolerance [@moller2009successful; @maklakov2011brains; @moller2015brain]. This finding highlights the importance of investigating trait-urbanization relationships in a phylogenetically informed manner.

@@

As illustrated above and in Appendix 1, many traits have been used to identify a species' adaptability to urban environments, but results have frequently differed among studies. As previously noted, a major shortcoming of previous studies has been the focus on discrete classifications of species based on their responses to urbanization. Our models assessed the applicability of traits in relation to one another (i.e., we did not run separate models or statistical tests for each specific trait as previously done; e.g., [@kark2007living]). The aforementioned differences in our results compared with others is likely due to increased sample size, as well as our use of a continuous measure of urbanization. Our broad-ranging analysis was made possible through the integration of broad-scale empirical datasets, which are rapidly expanding our knowledge of the impacts of urbanization on birds [e.g., @aronson2014global; @aronson2016hierarchical; @gutierrez2018citizen]. This is largely due to the growing use of citizen science data [@silvertown2009new] in combination with advances in remote sensing technology [@gorelick2017google]. We did not explicitly account for changes throughout the annual cycle in a species' association with urban environments [e.g., @la2014role; @lepczyk2017global; @katuwaleffect], but migratory status was considered as a predictor. We emphasize that future studies, especially within migrant-dominated systems (e.g., Europe, North America), should account for changes in a species' response to urbanization throughout the full-annual cycle [@hostetler2015full; @marra2015call]. Further, future studies should look to expand this analysis to other continents, and also investigate whether these patterns hold true at multiple spatial scales [@suarez2018scale] (i.e., does city size affect the patterns found here; [@sol2017urbanisation; @batary2018non]). Fully understanding the suite of traits which are associated with urban-tolerant species, and placing them in a theoretical landscape ecology context will ultimately allow for a better understanding of how to best preserve and maintain avian biodiversity in urban environments [@evans2011makes]. We highlight that future studies investigating the trait-urbanization relationship should focus on broad-scale data to form generalizable patterns which rely on continuous measures for improved predictive performance. This work should be further extended to test whether our findings can accurately predict species which are commonly found in urban environments in other parts of the world, based on a known suite of traits associated with a given set of species.

@@

```{r Figure 1, echo=FALSE, fig.cap="An example of five different species' average radiance value distributions, on a logarithmic scale. These species were chosen because they represent species with differnt distributional responses."}
knitr::include_graphics("finalFigs/Figure1.pdf")
```

@@

```{r Figure 3, echo=FALSE, fig.cap="A phylogenetic tree, based on Jetz et al. (2012) of the 477 bird species in the analysis. The tree demonstrates the urbanization index for each species in two ways: (1) coloring of the tips where dark red equates to a low urbanization index and dark blue equates to a high urbanization index, and (2) the size of the barplot portrayed for a species at its tip show the relative urbanization index. The urbanization index was taken as the median of the distribution for a species' average radiance values. Figure S2 shows a reference phylogenetic tree with the names of each tip displayed, and this link () provides an interactive visualization in order to easily identify each of the 477 species in the tree. For a full list of the 477 species included and their urbanization index see Appendix 4."}
knitr::include_graphics("finalFigs/Figure3.pdf")
```

@@

```{r Figure S1, echo=FALSE, fig.cap="A correlation plot for the continuous variables used in the analysis."}
knitr::include_graphics("finalFigs/FigureS1.pdf")
```

```{r Figure S2, echo=FALSE, fig.cap="A reference phylogenetic tree, with labeled tips, for each of the 477 species included in the analysis. Appendix 4 lists the 477 species."}
knitr::include_graphics("finalFigs/FigureS2.pdf")
```

```{r Figure S3, echo=FALSE, fig.cap="An assessment of phylogenetic uncertainty for some of the most important predictor variables included in the analysis."}
knitr::include_graphics("finalFigs/FigureS3.pdf")
```

>>>>

We filtered the bird observations [@callaghan2017assessing; @la2014role] by only including observations which were on complete checklists and which followed these protocols: travelling, random, stationary, area, or BirdLife Australia protocols [@barrett2003new]. For further information on eBird protocols see here: http://help.ebird.org/customer/portal/articles/1006209-understanding-observation-types. Further, we filtered the checklists to those with a distance travelled of < 5 km or an area covered which was < 500 Ha. Any checklists which were shared among multiple individuals were subsampled to include only one checklist in order to avoid duplicate results. After this sampling, we were left with 637,482 eligible checklists throughout Australia. Seabirds (gannets, petrels, shearwaters, etc.) were omitted from the analyses as we did not expect any correlation with seabirds using urban areas under normal circumstances. Only species which included > 100 observations in the eBird database over the specified time-frame, and that met our criteria were considered for further analyses. This left us with 580 potential species from the eBird dataset which had > 100 observations on eligible checklists in the specified time frame, which was a substantial proportion of the ~730 bird species which are resident in Australia. Given the diverse datasets (see below) used in this analysis, there were differing taxonomic authorities followed (see Appendix 2 for elaboration).

## Urbanization index
We used VIIRS night-time lights [@elvidge2017viirs] as a proxy of a bird's association with urban habitat. The VIIRS night-time lights measures the electric lighting present on Earth's surface which is generally associated with human settlement (Fig. S1). There are also a significant number of steps (e.g., filtering out background noise, degraded data, and other light source contamination) which are used to ensure the validity of the data and its association with human settlement [@elvidge2017viirs]. For the 637,482 checklists which met our criteria above, we used the checklists' associated spatial coordinates and Google Earth Engine [@gorelick2017google] to assign the average radiance value calculated from the VIIRS layer. For each checklist, the average radiance was calculated within a 5 km buffer, in order to match the spatial scale of the checklists included and to minimize any potential bias in eBird sampling protocols. Any radiance values which were negative were forced to 0.00001 [*sensu* @ou2015evaluation]. Each checklist's average radiance value was then passed to every bird on that checklist, which provided every species with a distribution of average radiance values (Fig. 1). The median for each species' given distribution of average radiance values served as the response variable in our analyses.

@@

Before modelling, the predictor variables were investigated for multi-collinearity. We used the corrplot package [@corrplot2017] to show correlation among continuous variables (Fig. S2), and used generalized variance inflation factors [@fox1992generalized], from the car package [@fox2011companion], which accounts for the degrees of freedom for a particular term to show minimal correlation amongst each other (Supplementary Table 1).

@@

### Phylogenetic analysis
For the phylogenetic portion of our analyses, we used the phylogenetic trees provided by Jetz et al. [@jetz2012global]. This phylogeny represents the first completely mapped phylogeny of 9,993 birds [@jetz2012global]. In order to assess whether there was a phylogenetic signal [@revell2008phylogenetic; @munkemuller2012measure] of the urbanization index and the continuous predictor variables used in our study, we used the phylosignal package [@keck2016phylosignal] in the R computing environment [@r2017coreteam]. This was also conducted for each of the continuous response variables in order to assess the relative contribution of phylogeny on the predictor variables.

@@

# RESULTS
A total of 316,306 checklists resulting in 5,944,819 observations of 477 different species were included in analyses. The least often recorded species was *Turnix melanogaster* (Black-breasted Buttonquail; N=104) while the most recorded species was *Cracticus tibicen* (Australian Magpie; N=158,615). The mean number of observations for a species was 10,250. Plots of all 477 species' distributions of average radiance values are presented in Appendix 3. A complete list of the 477 species, in addition to their traits, number of observations, and their urbanization indices can be found in Appendix 4.

## Urbanization index
For the 477 species included in the models the mean $\pm$ SD urbanization index was 0.91 $\pm$ 1.76, with a range from 0.008 (*Cracticus mentalis* Black-backed Butcherbird) to 11.85 (*Ninox strenua* Powerful Owl; Fig. 2).

### Phylogenetic signal
There was a distinct phylogenetic signal in our urbanization index measure (Fig. 3; Fig. S3; Table 1), as it was significantly different than random for 4 out of the 5 indices considered. Additionally, there was strong phylogenetic signal for potential predictor traits, especially body size and brain residual (Table 1). The result for body size ($k=3.3$) is among the strongest ever recorded measures of phylogenetic signal in the literature, and for $k$ the statistic suggests much stronger signal than expected from Brownian motion. We generally found weaker signal for behavioral traits compared with morphological traits. For these traits, the estimates of phylogenetic signal were weaker than Brownian motion but significant in most cases (Table 1).

@@

# DISCUSSION
Niche breadth, as measured by different types of generalism, was a significantly important life-history characteristic associated with the most urban-tolerant species in our analyses. This result was robust across non-phylogenetic and phylogenetic models, model-averaging approaches, and across phylogenetic uncertainty (Fig. S4). Conversely, specialized feeding strategies were negatively associated with urbanization: granivorous and insectivorous specialists were much less likely to be found in urban environments. These results suggest that the process of urbanization is, in general, operating to exclude specialist species [@sorace2009distribution]. Given specialist species are known to have strong negative responses to landscape fragmentation and disturbance [@devictor2008distribution], sustainable urban development with a goal of maintaining biological diversity in urban areas should focus on providing intact vegetation [@sushinsky2013should] connected by corridors [@litteral2017role]. A species-specific approach, which assesses particular resources necessary for specialist species, for example focusing on foraging [@shochat2004linking] and nesting [@fernandez2001habitat] requirements, is likely necessary. We recommend that a focal point of avian biodiversity conservation should be on increasing insect abundance and diversity in urban environments [@baldock2015uk], especially important given the recent noted collapse of insects in other parts of the world [@hallmann2017more].

The non-random distribution of the urbanization index across the phylogeny (Table 1) is consistent with the observation of reduced phylogenetic diversity in urban areas [@la2018phylogenetic], suggesting that particular phylogenetic constraints allow certain species to remain tolerant of urban environments [@morelli2016evidence; @ibanez2017global; @sol2017urbanisation]. This finding of a strong phylogenetic signal contrasts with previous studies [@evans2011makes], but this is likely a result of increased statistical power from our larger sample size: 477 species in this study compared with 55 species in Evans *et al.* [-@evans2011makes]. Unsurprisingly, our larger sample size found a non-random pattern in urbanization across the bird phylogeny and we further demonstrated the importance of treating species as non-independent by considering phylogeny in the context of the linear models. For example, some predictor variables became less important in the phylogenetic models compared with the non-phylogenetic models, including some feeding types (i.e., insect specialists) and habitat preferences (i.e., agricultural). For these variables, it is likely that there may be particular clades which include a number of species that are associated with both that particular trait and degree of urbanization [see @felsenstein1985confidence]. Conversely, some of the most important predictor variables (e.g., clutch size and feeding habitat generalism; Figures 4 and 5) showed a stronger relationship (i.e., increased parameter estimates) with urbanization when phylogeny was accounted for, and this affect was robust relative to a consideration of phylogenetic uncertainty (Fig. S4). Furthermore, residual brain mass (and body size to a lesser extent) was significantly important within the phylogenetic models, but not in the non-phylogenetic models, suggesting that within a clade the species with the largest residual brain mass [@maklakov2011brains] and largest body size [@major2010museum] are most suited to surviving in urban areas. While some studies have found a lack of evidence that residual brain size is an important predictor in urban-tolerant species [@kark2007living; @evans2011makes], our results suggest that residual brain size is indeed heavily correlated with urban-tolerance [@moller2009successful; @maklakov2011brains; @moller2015brain]. This finding highlights the importance of investigating trait-urbanization relationships in a phylogenetically informed manner.

@@

As illustrated above and in Appendix 1, many traits have been used to identify a species' adaptability to urban environments, but results have frequently differed among studies. As previously noted, a major shortcoming of previous studies has been the focus on discrete classifications of species based on their responses to urbanization. Our models assessed the applicability of traits in relation to one another [i.e., we did not run separate models or statistical tests for each specific trait as previously done; e.g., @kark2007living]. The aforementioned differences in our results compared with others is likely due to increased sample size, as well as our use of a continuous measure of urbanization. Our broad-ranging analysis was made possible through the integration of broad-scale empirical datasets, which are rapidly expanding our knowledge of the impacts of urbanization on birds [e.g., @aronson2014global; @aronson2016hierarchical; @gutierrez2018citizen]. This is largely due to the growing use of citizen science data [@silvertown2009new] in combination with advances in remote sensing technology [@gorelick2017google]. We did not explicitly account for changes throughout the annual cycle in a species' association with urban environments [e.g., @la2014role; @lepczyk2017global; @katuwaleffect], but migratory status was considered as a predictor. We emphasize that future studies, especially within migrant-dominated systems (e.g., Europe, North America), should account for changes in a species' response to urbanization throughout the full-annual cycle [@hostetler2015full; @marra2015call]. Further, future studies should look to expand this analysis to other continents, and also investigate whether these patterns hold true at multiple spatial scales [@suarez2018scale], by investigating the effect of city size [@sol2017urbanisation; @batary2018non] on these patterns. Fully understanding the suite of traits which are associated with urban-tolerant species, and placing them in a theoretical landscape ecology context will ultimately allow for a better understanding of how to best preserve and maintain avian biodiversity in urban environments [@evans2011makes]. We highlight that future studies investigating the trait-urbanization relationship should focus on broad-scale data to form generalizable patterns which rely on continuous measures for improved predictive performance. This work should be further extended to test whether our findings can accurately predict species which are commonly found in urban environments in other parts of the world, based on a known suite of traits associated with a given set of species.

@@

```{r Figure 1, echo=FALSE, fig.cap="An example of five different species' average radiance value distributions, on a logarithmic scale. These species were chosen because they represent species with different distributional responses."}
knitr::include_graphics("finalFigs/Figure1.pdf")
```

@@

```{r Figure 3, echo=FALSE, fig.cap="A phylogenetic tree, based on Jetz et al. (2012) of the 477 bird species in the analysis, demonstrating the clustering of the urbanization index in regards to phylogeny. The tree demonstrates the urbanization index for each species in two ways: (1) coloring of the tips where dark red equates to a low urbanization index and dark blue equates to a high urbanization index, and (2) the size of the barplot portrayed for a species at its tip show the relative urbanization index. The urbanization index was taken as the median of the distribution for a species' average radiance values. Figure S3 shows a reference phylogenetic tree with the names of each tip displayed, and this link () provides an interactive visualization in order to easily identify each of the 477 species in the tree and their respective urbanization indices. For a full list of the 477 species included and their urbanization index see Appendix 4."}
knitr::include_graphics("finalFigs/Figure3.pdf")
```

@@

```{r Figure S1, echo=FALSE, fig.cap="The VIIRS nightime-lights shown for Australia, Brisbane, Melbourne, and Sydney. Values are measured as average radiance shown in red."}
knitr::include_graphics("finalFigs/FigureS1.pdf")
```

```{r Figure S2, echo=FALSE, fig.cap="A correlation plot for the continuous variables used in the analysis."}
knitr::include_graphics("finalFigs/FigureS2.pdf")
```

```{r Figure S3, echo=FALSE, fig.cap="A reference phylogenetic tree, with labeled tips, for each of the 477 species included in the analysis. Appendix 4 lists the 477 species."}
knitr::include_graphics("finalFigs/FigureS3.pdf")
```

```{r Figure S4, echo=FALSE, fig.cap="An assessment of phylogenetic uncertainty for some of the most important predictor variables included in the analysis."}
knitr::include_graphics("finalFigs/FigureS4.pdf")
```
